---
title: "A strategy to assess water meter performance"
output: github_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE, fig.align="center")
```



```{r, echo=FALSE}
library(forecast)
library(strucchange)
library(trend)
library(ggplot2)
```



```{r, echo=FALSE}
load("~/GitHub/StrategyWaterMeterPerformance/workspace_data_functions.RData")
```

## Overview

Functions, datasets and script used in the paper *A strategy to assess water meter performance* by Clara Cordeiro, Ana Borges and M. Rosario Ramos, in *Journal of Water Resources Planning and Management*, 2021. Status: *under revision*.

### Install packages

```R
install.packages("forecast",dependencies = TRUE)
install.packages("strucchange",dependencies = TRUE)
install.packages("trend",dependencies = TRUE)
```

### Usage 

```R
library(forecast)  
library(strucchange) 
library(trend) 
load("workspace_data_functions.RData")
```

### Functions implemented 

- **test.Outliers.STL**: *TRUE* in case of outliers, otherwise *FALSE*

- **T.dec**: *TRUE* in case of a decreasing trend using Mann-Kendall test, otherwise *FALSE*

- **s.bp**: find the relevant breakpoints

- **RelMChange**: calculate the RMC (Eq. 5)


### Example - Residential household 2 (RH2)

```R
y<-RH2
```

```{r fig.height=3,fig.width=6}
autoplot(RH2)+
  geom_line(na.rm=TRUE,color="darkorange",size=1)+
  geom_point()+
  #ylim(0,20)+
  theme_classic()+
  ggtitle("RH2") +
  xlab(" ") + ylab(expression(m^3))

```

#### Initial settings

```R
al<-0.05  # level of significance
s<-12   # monthly
n.y<-length(y)  
k<-3  # Mean Absolute Error, but it can be the Mean Error (k<-1), Root Mean Square Error (k<-2); see accuracy() from package forecast for more information about it.
```

#### Step 1 - Seasonal-trend decomposition based on Loess

```R
rob.y<-test.Outliers.STL(y) # check outliers
fit.y<-stl.fit(y,rob.y,k)  # stl.fit function
```

```{r fig.height=4,fig.width=6}
autoplot(fit.y$stlfit,range.bars=FALSE) +
  geom_line(na.rm=TRUE,color="darkorange",size=1)+
  theme_bw(base_size = 8)+
  ggtitle("RH2") +
  xlab(" ") +theme(title = element_text(size = 11))
```

#### Step 2 - Detecting *breakpoints* (bp)

```R
ystar<-y-fit.y$stlfit$time.series[,"seasonal"] # seasonally adjusted (y*)  (Eq.2)
min.h<-round(s/n.y,2) # max breaks (Eq. 3)
max.break<-ceiling(0.5/min.h) # min h (Eq. 4)
bpy<-breakpoints(ystar~1,h = min.h,breaks=max.break) # breakpoints() from package strucchange
```

```{r fig.height=3,fig.width=6}
p<-autoplot(ystar)+
  geom_line(na.rm=TRUE,color="darkorange",size=1)+
  theme_classic()+
  ylim(0,20)+
  xlab(" ") + ylab(expression(m^3))+
  ggtitle("RH2")

my.bp1<- breakdates(bpy)
p11<-p  + geom_vline(xintercept = my.bp1, size=1,col="grey", linetype = "dashed")
my.bp2 <- index(ystar)[bpy.star]
p2<-p11  + geom_vline(xintercept = my.bp2, size=1,col="grey", linetype = "dashed")
p2
```


#### Step 3 - Detecting *relevant breakpoints* (bp*, blue dashed vertical line below)

```R
bpy.star<-s.bp(ystar,bpy$breakpoints,al)
```
```{r fig.height=3,fig.width=6}
p<-autoplot(ystar)+
  geom_line(na.rm=TRUE,color="darkorange",size=1)+
  theme_classic()+
  ylim(0,20)+
  xlab(" ") + ylab(expression(m^3))+
  ggtitle("RH2")

my.bp1<- breakdates(bpy)
p11<-p  + geom_vline(xintercept = my.bp1, size=1,col="grey", linetype = "dashed")
my.bp2 <- index(ystar)[bpy.star]
p2<-p11  + geom_vline(xintercept = my.bp2, size=1,col="blue", linetype = "dashed")
p2
```

#### Step 4 - Determining *Relative Magnitude of Change* (RMC) 

If bp*=NULL then **stop** otherwise calculate

```R
rmc<-RelMChange(ystar,bpy$breakpoints,bpy.star) (Eq. 5)
```

#### Results 

Table 1 and table 2, separated by double **|  |**. 

```{r, echo=FALSE}
  
  aux<-bpy$breakpoints
  
  # compute breakpoint dates
  barray = c()
  for(a in aux){
    barray <- append(barray, paste0(floor(time(y)[a]),'(',month.abb[round(((time(y)[a] %% 1)*12) + 1)],')'))
    }
  
  # add beginning and end indices so we can loop over segments
  aux = c(1, aux, n.y)
  
  # save the breakdates to the output string
  output <- paste0(barray, collapse=',') 
  output <- paste0(output, ' |')
  
  output.seg<-NULL
  # loop over each segment and save the results to the output string
  for(ai in 1:(length(aux)-1)){
    
    seg<-trend::sens.slope(ystar[aux[ai]:aux[ai+1]])
    
    # append the segment data
    output.seg<-paste(output.seg,'Segment', ai ,' |')
    output<-paste(output, 
                   paste0(round(seg[["estimates"]][["Sen's slope"]], 2), '(',round(seg[["p.value"]], 5),')', ' |'))
    }
  
 ### table 2 (add to table 1)
  
  # step3: select bp*
  bpy.star<-s.bp(ystar,bpy$breakpoints,al)
  
  
  if(is.null(bpy.star)) {
    output2<-paste("No relevant breakpoints", ' |')
} else {
  aux2<-bpy.star
  
  # compute breakpoint dates
  barray = c()
  for(a in aux2){
    barray <- append(barray, paste0(floor(time(y)[a]),'(', month.abb[round(((time(y)[a] %% 1)*12) + 1)], ')'))
    }
  
  # save the breakdates to the output string
  output2 <- paste0(barray, collapse=',') 

  # step4: calculate RMC  Eq. (5)
  # if bp* NULL then stop otherwise do the following
  ## magnitude
  
  rmc<-RelMChange(ystar,bpy$breakpoints,bpy.star)
  
  output.rmc<-c()
  for(ai in 1:length(rmc)){
    output.rmc<-paste(output.rmc, 
                  paste0(round(rmc[ai], 2), ' |'))
  }
  output2<-paste(output2, ' |', output.rmc)
  }
  
  
print(paste('Type | Breakpoints |',output.seg, '| Relevant bp* | RMC '))
  
print(paste('Rh2 |', output, ' | ', output2))
```

Run *script_run_data.R* for all data in the paper.

### About software used

- **R** version 4.1.0
- package **trend** version 1.1.4
- package **forecast** version 8.14
- package **strucchange** version 1.5-2

#### *Contributions, issues, and feature requests are welcome!*